Short-term synaptic and neuronal plasticity act in synergy for deviance detection in spiking networks

### 2022-12-13 paper-1/manuscript.tex
Tweak Fig4 (mean, not med, dV_TA)

We quantified the former with the strength of adaptation $V_{TA}$ (Equation \ref{eqn-TA}), i.e., the amount by which the firing threshold was raised due to TA. As shown in Figure \ref{fig4}b, there was a clear and widespread lowering of thresholds at the start of dev trials relative to msc, visible both in the raw data (left and middle panels) and in the contrast (right panel, early blue portions). In contrast, once the bulk of activity had occurred, many neurons showed higher thresholds in dev (right panel, later orange portions), since they spiked more often. Finally, note that inhibitory neurons, which were modeled without TA, naturally showed no difference in $V_{TA}$.

@@

\begin{figure*}%
    \centering
    \includegraphics{fig4.pdf}
    \caption{}
    \label{fig4}
\end{figure*}
\textbf{Fig. 4} Responses and threshold adaptation with TA only.
\textbf{a} Target trial average post-stimulus spike histogram, showing trial time along the horizontal axis, and neurons along the vertical, sorted by the time of the first recorded spike across all trials and conditions. Left and middle column, target trials in dev sequence and msc sequence, respectively; right column, contrast between these two.
\textbf{b} Target trial average of the threshold adaptation voltage $V_{TA}$, with axes, neuron order and columns as in panel a.
\textbf{c} Target trial average of the membrane potential $V_m$, with axes, neuron order and columns as in panel a.
\textbf{d} Relationship across excitatory neurons between the contrast (dev - msc) in $V_{TA}$ at the start of target trials, and the contrast in average number of spikes fired in target trials, and associated linear regression.
\textbf{e} Median contrast (dev - msc) in $V_{TA}$ across networks and stimuli, plotted against the Pearson correlation coefficients $\rho$ of the relationship shown in the previous panel ($\\Delta V_{TA}$ vs response size contrast across excitatory neurons). Each marker corresponds to one network and target stimulus. Red markers indicate a significant correlation (p < 0.05). Inset histograms show the distribution of the corresponding values, with orange portions indicating the 95\% confidence interval of the median, calculated by bootstrap with 10000 iterations.
\textbf{f} In the sample network, contribution of the contrast in $V_{TA}$ and $V_m$ to bins with increased firing in dev (i.e., positive/red bins in the spike probability contrast, panel a), weighted by that same contrast.
\textbf{g} Contribution of $V_{TA}$ relative to the sum of the absolute values of the weighted contributions in panel f.
\textbf{h} Relative contribution of $V_{TA}$ across networks and stimuli, showing median (solid line) and inter-quartile range (shaded area)

Based on these measures, we then asked why there were more spikes in dev trials. We first focused on the contribution of $V_{TA}$ alone in an effort to confirm that thresholds are lower in dev, and that this reduction in thresholds causes increased firing. We correlated the level of $V_{TA}$ at the start of target trials (i.e., at the left edge of the histograms in Figure \ref{fig4}b) with the average target trial response of each excitatory neuron in terms of spikes fired. In the sample network (Figure \ref{fig4}d), we found that thresholds were clearly lower in dev than msc (t = -17.3, p = 2.57e-57, median -0.166 mV; one-sided t-test, n=800), and this decrease was accompanied by an increase in the target trial response at the level of individual neurons (Pearson correlation, $\rho$ = -0.424, p = 3.1e-36). As shown in Figure \ref{fig4}e, this relationship held across networks and stimuli: Median $V_{TA}$ was lower in dev than msc (t = -11.2, p = 1.08e-20, median = -0.094 mV; one-sided t-test, n=120), and the neuron-level correlation between the contrasts in $V_{TA}$ and target response was almost universally negative (t = -23.1, p = 3.72e-46, median = -0.357). This indicates that networks were less adapted during oddball sequences, and that this lack of adaptation played an important role in the increased response to deviant inputs.

>>>>

We quantified the former with the strength of adaptation $V_{TA}$ (Equation \ref{eqn-TA}), i.e., the amount by which the firing threshold was raised due to TA. As shown in Figure \ref{fig4}b, there was a clear and widespread lowering of thresholds at the start of target dev trials relative to msc, visible both in the raw data (left and middle panels) and in the contrast (right panel, early blue portions). In contrast, once the bulk of activity had occurred, many neurons showed higher thresholds in dev (right panel, later orange portions), since they spiked more often. Finally, note that inhibitory neurons, which were modeled without TA, naturally showed no difference in $V_{TA}$.

@@

\begin{figure*}%
    \centering
    \includegraphics{fig4.pdf}
    \caption{}
    \label{fig4}
\end{figure*}
\textbf{Fig. 4} Responses and threshold adaptation with TA only.
\textbf{a} Target trial average post-stimulus spike histogram, showing trial time along the horizontal axis, and neurons along the vertical, sorted by the time of the first recorded spike across all trials and conditions. Left and middle column, target trials in dev sequence and msc sequence, respectively; right column, contrast between these two.
\textbf{b} Target trial average of the threshold adaptation voltage $V_{TA}$, with axes, neuron order and columns as in panel a.
\textbf{c} Target trial average of the membrane potential $V_m$, with axes, neuron order and columns as in panel a. The first 10 ms of the neurons under direct stimulation are masked out to avoid saturation of the color mask.
\textbf{d} Relationship across excitatory neurons between the contrast (dev - msc) in $V_{TA}$ at the start of target trials, and the contrast in average number of spikes fired in target trials, and associated linear regression.
\textbf{e} Mean contrast (dev - msc) in $V_{TA}$ across networks and stimuli, plotted against the Pearson correlation coefficients $\rho$ of the relationship shown in the previous panel ($\\Delta V_{TA}$ vs response size contrast across excitatory neurons). Each marker corresponds to one network and target stimulus. Red markers indicate a significant correlation (p < 0.05). Inset histograms show the distribution of the corresponding values, with orange portions indicating the 95\% confidence interval of the median, calculated by bootstrap with 10000 iterations.
\textbf{f} In the sample network, contribution of the contrast in $V_{TA}$ and $V_m$ to bins with increased firing in dev (i.e., positive/red bins in the spike probability contrast, panel a), weighted by that same contrast.
\textbf{g} Contribution of $V_{TA}$ relative to the sum of the absolute values of the weighted contributions in panel f.
\textbf{h} Relative contribution of $V_{TA}$ across networks and stimuli, showing median (solid line) and inter-quartile range (shaded area)

Based on these measures, we then asked why there were more spikes in dev trials. We first focused on the contribution of $V_{TA}$ alone in an effort to confirm that thresholds are lower in dev, and that this reduction in thresholds causes increased firing. We correlated the level of $V_{TA}$ at the start of target trials (i.e., at the left edge of the histograms in Figure \ref{fig4}b) with the average target trial response of each excitatory neuron in terms of spikes fired. In the sample network (Figure \ref{fig4}d), we found that thresholds were clearly lower in dev than msc (t = -17.3, p = 2.57e-57, median -0.166 mV; one-sided t-test, n=800), and this decrease was accompanied by an increase in the target trial response at the level of individual neurons (Pearson's $\rho$ = -0.424, p = 3.1e-36). As shown in Figure \ref{fig4}e, this relationship held across networks and stimuli: Average $V_{TA}$ was lower in dev than msc (t = -7.32, p = 1.58e-11, median = -0.074 mV; one-sided t-test, n=120), and the neuron-level correlation between the contrasts in $V_{TA}$ and target response was almost universally negative (t = -23.1, p = 3.72e-46, median = -0.357). This indicates that networks were less adapted during oddball sequences, and that this lack of adaptation played an important role in the increased response to deviant inputs.
